Viscous Transport in Eroding Porous Media

### 2019-10-02 docs/transport/erosion_transport.tex
NM

In Figure~\ref{fig:Eroding20Transport}(a), we plot the tortuosity as a
function of the porosity. The initial porosity is $\phi = 37.68\%$, and
the initial tortuosity is $T = 1.16$.  The tortuosity is computed with both
the length of the streamlines~\eqref{eqn:tortuosity1} (red stars) and
using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in Figure~\ref{fig:Eroding20tort}(c).
The two tortuosity formulas give similar results, and any discrepancy
can be accounted for by slow regions of recirculation and from applying
quadrature to compute the tortuosity.  As the bodies erode, wide
channels form where streamlines undergo only minor vertical deflections,
and this explains why the tortuosity eventually decreases with porosity.
We computed lines of best fit using the porosity-tortuosity
models~\eqref{eqn:tortuosityModels}, and the power law model gave the
smallest error.  The  black dashed line in
Figure~\ref{fig:Eroding20Transport}(a) is the line of best fit
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.2064}
\end{align}
with a root-mean-square error of $5.90 \times 10^{-3}$.  Interestingly,
at the low porosities, the tortuosity initially increases. This increase
occurs because in the absence of erosion (left plot in
Figure~\ref{fig:Eroding20vort}), many of the streamlines, such as those
initialized between bodies 15 \& 18, only perform minor deflections to
pass through the narrow regions, albeit, very slowly.  However, as
erosion starts to open the channels, the streamlines deflect into the
fast regions, such as the region above body 11, and this increases the
amount of vertical deflection, and therefore the tortuosity.

In Figure~\ref{fig:Eroding20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  The spreading is
computed at seven geometries of different porosities that are formed
during the erosion process.  So that the spreading reaches a statistical
equilibrium, we use the reinsertion algorithm described in
Section~\ref{sec:dispersion} to form sufficiently long trajectories.
For all the reported porosities, the particle dispersion exhibits two
power low regimes.  Initially, the dispersion is ballistic
($\sigma_\lambda \sim t$) since individual fluid particles have not yet
explored enough space to significantly alter their velocity.  However,
once the particles have been subjected to a range of velocities, their
dispersion slows down, and we observe super-dispersive ($\sigma_\lambda
\sim t^\alpha$, $\alpha \in (1/2,1)$) behavior over at least one order
of magnitude in time.  Before any erosion takes place, the anomalous
dispersion coefficient is $\alpha = 0.56$.  Then, as the grains begin to
erode, the dispersion rate grows towards the ballistic regime that
occurs in the absence of grains.  The monotonic increase in dispersion
with respect to the porosity is explained by the onset of channels where
many tracers experience less variability in their velocities.

>>>>

In Figure~\ref{fig:Eroding20Transport}(a), we plot the tortuosity as a
function of the porosity. The initial porosity is $\phi = 37.68\%$, and
the initial tortuosity is $T = 1.16$.  The tortuosity is computed with both
the length of the streamlines~\eqref{eqn:tortuosity1} (red stars) and
using the spatial average of the velocity on an Eulerian
grid~\eqref{eqn:tortuosity2} (blue marks).  The red square corresponds
to the porosity of the geometry in Figure~\ref{fig:Eroding20tort}(c).
The two tortuosity formulas give similar results, and any discrepancy
can be accounted for by slow regions of recirculation and from applying
quadrature to compute the tortuosity.  As the bodies erode, wide
channels form where streamlines undergo only minor vertical deflections,
and this explains why the tortuosity eventually decreases with porosity.
We computed lines of best fit using the porosity-tortuosity
models~\eqref{eqn:tortuosityModels} and \nick{found the power law to give} the
smallest error.  The  black dashed line in
Figure~\ref{fig:Eroding20Transport}(a) is the line of best fit
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.2064}
\end{align}
with a root-mean-square error of $5.90 \times 10^{-3}$.  Interestingly,
at the low porosities, the tortuosity initially increases. This increase
occurs because in the absence of erosion (left plot in
Figure~\ref{fig:Eroding20vort}), many of the streamlines, such as those
initialized between bodies 15 \& 18, only perform minor deflections to
pass through the narrow regions, albeit, very slowly.  However, as
erosion starts to open the channels, the streamlines deflect into the
fast regions, such as the region above body 11, and this increases the
amount of vertical deflection, and therefore the tortuosity.

\todo[inline]{I wonder if this initial increase in the tortuosity is specific to the process of erosion, or if it would also occur if we simply took a collection of circular bodies and changed the porosity by controlling the radii? We don't have to answer this question now, but it would be interesting to consider in the future.}

In Figure~\ref{fig:Eroding20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  The spreading is
computed \nick{\st{at} for} seven geometries of different porosities that are formed
during the erosion process.  So that the spreading reaches a statistical
equilibrium, we use the reinsertion algorithm described in
Section~\ref{sec:dispersion} to form sufficiently long trajectories.
For all the reported porosities, the particle dispersion exhibits two \nick{distinct
power \st{low} law} regimes.  Initially, the dispersion is ballistic
($\sigma_\lambda \sim t$) since individual fluid particles have not yet
explored enough space to significantly alter their velocity.  However,
once the particles have been subjected to a range of velocities, their
dispersion slows \nick{\st{down}}, and we observe super-dispersive ($\sigma_\lambda
\sim t^\alpha$, $\alpha \in (1/2,1)$) behavior over at least one order
of magnitude in time.  Before any erosion takes place, the anomalous
dispersion coefficient is $\alpha = 0.56$.  Then, as the grains begin to
erode, the dispersion rate grows towards the ballistic regime that
occurs in the absence of grains.  The monotonic increase in dispersion
with respect to the porosity is explained by the onset of channels where
many tracers experience less variability in their velocities.
